---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ecohabr2

<!-- badges: start -->
<!-- badges: end -->

The goal of ecohabr2 is to analyze raw data from the Eco-HAB experiments.

Eco-HAB is an automated system for ecologically relevant assessment of social behaviors in mice. For detailed information please see Puścian et. al. (https://doi.org/10.7554/eLife.19532)

ecohabr2 is an independent R implementation of analytic methods in the above and other publications using Eco-HAB. Care is taken to reproduce the analysis. However, due to possible difference in raw data cleansing, algorithms and computational precision, the results may not be exactly
the same as the original methods.

## Installation

You can install the development version of ecohabr2 from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("horizon-mm/ecohabr2")
```

## Example

This is a basic example which shows you how to use this package:

```{r example, eval=FALSE}
library(ecohabr2)
# Define the files for mouse ID, experimental time setting and raw data files.
# Be sure to choose "idfile" or "timefile" first, otherwise choose.dir() may fail in RStudio.
idfile <- file.choose()
timefile <- file.choose()
files <- list.files(choose.dir(), pattern = "0000\\.txt$", full.names = TRUE)
# Create an EcoHAB object and calculate cage/tube travelling events.
ehd <- EcoHAB$new(files, idfile, timefile)
ehd$calc_events()
# Optional: some RFID reading may be missing and cause abnormal data (extreme long tube events),
# which can be manually viewed and adjusted. 
ehd$edit_events()
# Calculate social events
ehd$calc_single()
ehd$calc_pair()
ehd$calc_follow()
# Export events list as csv files.
cohort <- "my_cohort" # Use your own cohort name
for (name in ehd$all_results())
  data.table::fwrite(ehd$get_result(name), paste(cohort, name, "csv", sep = "."))
# For certain time bins, calculate activities in terms of duration and counts,
# and export results as csv files.
for (binsize in c(3600, 14400, 43200)) {
  tag <- paste0(binsize/3600, "hr")
  ehd$set_binsize(binsize)
  ehd$calc_activity("events")
  ehd$calc_activity("single")
  ehd$calc_activity("pair")
  ehd$calc_activity("follow")
  for (name in ehd$all_results()) {
    if (grepl("time$", name) || grepl("visits$", name))
      data.table::fwrite(ehd$get_result(name), paste(cohort, name, tag, "csv", sep = "."))
  }
}
```
